---
title: "Is Stop-and-Frisk Racially Motivated?"
subtitle: "A Closer Look into the NYPD’s Practices"
author: "Arielle Barylsky, Sophia Foster, Rachel Yan"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    highlight: tango
    theme: cosmo
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: kable

---



```{r include=FALSE}
#Load necessary packages
library(readr)
library(plyr)
library(stringr)
library(tidyverse)
library(RColorBrewer)
```

```{r include=FALSE}

#Load 2010 demographic dataset

dem2010 <- read_csv("ACS_10_5YR_DP05.csv",
                    skip = 1)
                    
#Select only the needed columns
dem2010 <- dem2010 %>%
  select(-contains("Margin of Error")) %>%
  select(3, contains("Percent")) %>%
  select(c(1, 2, 3, 4, 5, 6, 7, 8))
  
#Rename columns
names(dem2010)[1] <- "Borough"
names(dem2010)[2] <- "Total"
names(dem2010)[3] <- "White"
names(dem2010)[4] <- "Black"
names(dem2010)[5] <- "American Indian and Alaska Native"
names(dem2010)[6] <- "Asian"
names(dem2010)[7] <- "Native Hawaiian and Other Pacific Islander"
names(dem2010)[8] <- "Other"

#Combine Asian and Native Hawaiian and Other Pacific Islander
dem2010 <- dem2010 %>%
  mutate(`Asian and Pacific Islander` = `Asian` + `Native Hawaiian and Other Pacific Islander`)

#Remove unnecessary columns
dem2010$`Asian` <- NULL
dem2010$`Native Hawaiian and Other Pacific Islander` <- NULL

#Tidy the columns 
dem2010 <- dem2010 %>%
  gather(key = "Race",
  value = "Percent",
  "White":"Asian and Pacific Islander")
  
#Add Year column
dem2010$Year <- rep(2010, nrow(dem2010))
```

```{r include=FALSE}
#Load 2017 demographic dataset
dem2017 <- read_csv("ACS_17_5YR_DP05.csv",
                    skip = 1)
                    
#Select only the needed columns
dem2017 <- dem2017 %>%
  select(c(3, 20, 26, 30, 34, 38, 42, 46))

#Rename columns
names(dem2017)[1] <- "Borough"
names(dem2017)[2] <- "Total"
names(dem2017)[3] <- "White"
names(dem2017)[4] <- "Black"
names(dem2017)[5] <- "American Indian and Alaska Native"
names(dem2017)[6] <- "Asian"
names(dem2017)[7] <- "Native Hawaiian and Other Pacific Islander"
names(dem2017)[8] <- "Other"

#Combine Asian and Native Hawaiian and Other Pacific Islander
dem2017 <- dem2017 %>%
  mutate(`Asian and Pacific Islander` = `Asian` + `Native Hawaiian and Other Pacific Islander`)

#Remove unnecessary columns
dem2017$`Asian` <- NULL
dem2017$`Native Hawaiian and Other Pacific Islander` <- NULL

#Tidy the columns 
dem2017 <- dem2017 %>%
  gather(key = "Race",
  value = "Percent",
  "White":"Asian and Pacific Islander")
  
#Add Year column
dem2017$Year <- rep(2017, nrow(dem2017))
```

```{r include=FALSE}
#Combine 2010 and 2017 cleaned demographic datasets
fulldem <- rbind(dem2010, dem2017)

#Rename values in "Borough" from County names to Borough names
fulldem$Borough <-
  revalue(
    fulldem$Borough,
    c(
      "Bronx County, New York" = "The Bronx",
      "Kings County, New York" = "Brooklyn",
      "New York County, New York" = "Manhattan",
      "Queens County, New York" = "Queens",
      "Richmond County, New York" = "Staten Island"
    )
  )
```

# What is stop-and-frisk (and why do we care)? 

Although laws vary by state, the basic idea of stop-and-frisk is that to legally stop someone, an officer “must have a reasonable suspicion that a crime has been, is being, or is about to be committed by the suspect.”^[Busby, John C. “Stop and Frisk.” LII / Legal Information Institute, 17 Sept. 2009, https://www.law.cornell.edu/wex/stop_and_frisk.]  The New York Police Department has cataloged stop-and-frisk data for the past 14 years in the five boroughs of New York City, from 2003 to 2017.^[“Stop, Question and Frisk Data.” NYPD. https://www1.nyc.gov/site/nypd/stats/reports-analysis/stopfrisk.page. Accessed 15 Dec. 2018]  Considering the history of racism in the United States and the myriad of news surrounding racially-motivated police brutality, we decided to examine this practice for racial bias among people frisked, searched, and subject to physical force. 


To start, the graph below visualizes the racial makeup of the five boroughs of New York City. To have a baseline for examining potential racial bias, we needed race demographic data for each borough. We found population demographics data from the ACS Census 5-Year Estimates and selected the years 2017 and 2010 to compare with the NYPD data. 

```{r}
#Change order of race levels for graph
race_levels <-
  c(
  "American Indian and Alaska Native",
  "Asian and Pacific Islander",
  "Black",
  "White",
  "Other"
  )
  
fulldem <- fulldem %>%
  mutate(Race = factor(Race, levels = race_levels))
  
#Change order of borough levels for graph
borough_levels <-
  c("The Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island")
  
fulldem <- fulldem %>%
  mutate(Borough = factor(Borough, levels = borough_levels))
```

```{r fig.width = 10}

#Plot race demographics in each borough for both years
ggplot(data = fulldem,
       mapping = aes(x = Year, y = Percent, fill = Race)) + geom_col(position =
       position_dodge()) + facet_wrap(~ Borough) + scale_x_continuous(breaks =
       c(2010, 2017)) + scale_fill_brewer(palette = "Set1") +
       ggtitle("Racial Distribution by Borough")
```

The racial makeup of the boroughs did not significantly change in the 7-year span. In all the boroughs, the Indigenous population makes up less than 1% of the population. White makes up the largest racial demographic on average (~50% in 2010), but is not evenly distributed across the boroughs. Black stands at an average of 24%, with the highest percentage in the Bronx (36%) and the lowest in Staten Island (11%).

If NYPD stop-and-frisk practices were not racially motivated, we would expect stop-and-frisk rates to follow about the same racial breakdown as the population demographics.
